From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study

### 2025-06-27 draft_v1/draft_v2.qmd
update draft_v2

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    template-partials:
      - title.tex
    include-in-header: header.tex
    crossref:
      appendix-title: "App."
      appendix-delim: ":"
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
nocite: |
  @*
csl: apa.csl
abstract: |
  \noindent This study investigates the dynamic relationship between precarious life conditions and depression by combining constraint-based causal discovery with computational modeling. Using cross-sectional data from the HELIUS study, we applied cycle-capable causal discovery algorithms to uncover directional relationships between domain-specific precariousness indicators, and depressive symptoms. Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—identified financial stress as a robust driver of depression, and highlighted key symptoms, such as sleep disturbance and depressed mood, as likely initiators or reinforcers of downstream precariousness.
  To explore how these structural insights shape intervention effectiveness, we developed a simplified nonlinear dynamical model simulating the co-evolution of depression and social precariousness under varying levels of financial stress reduction. Simulation results revealed that the system's responsiveness depends not only on the intensity of external intervention but also on internal structural features such as feedback strength and noise. In particular, some configurations exhibited bistability, requiring interventions to surpass a critical threshold to trigger lasting improvement, while others responded more proportionally.
  Together, these findings emphasize that internal feedback dynamics—not just external adversity—can shape the impact of mental health interventions. More broadly, our study offers a scalable framework for integrating causal discovery and dynamical simulation to better understand and inform policy responses to complex psychosocial phenomena.

@@

As illustrated in @fig-concept, the model includes three components: depression ($D$), represented by the PHQ-9 sum score, social precariousness ($P$), and external financial stress ($S$). Depression and precariousness are modeled as coupled state variables that influence each other over time, with $S$ acting as a shared input. These dynamics are formalized as a pair of nonlinear stochastic differential equations (SDEs):

$$
\begin{aligned}
dP &= \lambda_P \left( \tanh(\alpha_1 S + \alpha_2 D) - P \right) dt + \sigma_1 dW_1 \\
dD &= \lambda_D \left( \tanh(\beta_1 S + \beta_2 P) - D \right) dt + \sigma_2 dW_2 
\end{aligned}
$$

Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress. The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which variable adjusts toward its input-driven target state. The coefficients $\alpha_1, \alpha_2$ and $\beta_1, \beta_2$ represent the strengths of the causal influences between variables. 
Stochastic fluctuations are introduced through the terms $\sigma_1 dW_1$ and $\sigma_2 dW_2$, where $dW_1$ and $dW_2$ are independent Wiener processes. The system uses the hyperbolic tangent function ($tanh$) to bound the influence of inputs, ensuring smooth, saturating dynamics and preventing runaway feedback loops. Importantly, the nonlinearity introduced by the $tanh$ function also allows the system to exhibit bistability, supporting the existence of two stable states: one corresponding to low depression and precariousness, and the other to high depression and precariousness.

<!-- Building on the causal structure derived from the result of causal discovery, we implement a high-level simulation model to explore the potential impact of targeted interventions on precariousness. This serves as a proof-of-concept for translating causal relationships into dynamic systems that can generate actionable insights. -->

<!-- The structure of the simulation model, illustrated in @fig-concept, captures the dynamic interplay between depression ($PHQsum$) and social precariousness ($P.soc$), both of which are influenced by an external financial stressor ($S.fin$). Importantly, the model incorporates cyclic dependencies between depression and precariousness, reflecting their reciprocal influence over time. -->
<!-- This system is formalized as a set of non-linear stochastic differential equations (SDEs): -->

<!-- $$ -->
<!-- \begin{aligned} -->
<!-- dP &= \lambda_P \left( \tanh(\alpha_1 S + \alpha_2 D) - P \right) dt + \sigma_1 dW_1 \\ -->
<!-- dD &= \lambda_D \left( \tanh(\beta_1 S + \beta_2 P) - D \right) dt + \sigma_2 dW_2  -->
<!-- \end{aligned} -->
<!-- $$ -->

<!-- Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress. The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which variable adjusts toward its input-driven target state. The coefficients $\alpha_1, \alpha_2$ and $\beta_1, \beta_2$ represent the strengths of the causal influences between variables.  -->
<!-- Stochastic fluctuations are introduced through the terms $\sigma_1 dW_1$ and $\sigma_2 dW_2$, where $dW_1$ and $dW_2$ are independent Wiener processes. The system uses the hyperbolic tangent function ($tanh$) to bound the influence of inputs, ensuring smooth, saturating dynamics and preventing runaway feedback loops. Importantly, the nonlinearity introduced by the $tanh$ function also allows the system to exhibit bistability, supporting the existence of two stable states: one corresponding to low depression and precariousness, and the other to high depression and precariousness. -->

@@

### Model Calibration

To identify realistic parameter settings for the simulation model, we calibrate six free parameters—$\alpha_1$, $\alpha_2$, $\beta_1$, $\beta_2$, and noise amplitudes $\sigma_1$, $\sigma_2$—using the *Non-dominated Sorting Genetic Algorithm II* (NSGA-II) [@deb2002fast], a multi-objective evolutionary optimization method well-suited for noisy and high-dimensional search spaces. 
The goal is to match simulated system behavior to empirical summary statistics derived from the HELIUS dataset. Specifically, we aim to match the observed means and variances of depression ($D$) and social precariousness ($P$), along with all three partial correlations among $D$, $P$, and financial stress ($S$): $\rho_{DP.S}$, $\rho_{DS.P}$, and $\rho_{PS.D}$.

Given the stochastic nature of the model, each candidate parameter set is evaluated by simulating 1,000 virtual individuals, each assigned a fixed $S$ value sampled from the empirical distribution. From these simulations, we compute population-level summary statistics based on the final states. The fitness of each parameter set is measured by how closely its summary statistics matched the empirical data, using the total absolute deviation across all objectives.

<!-- NSGA-II is implemented via the `nsga2R` package [@nsga2R], with a population of 500 evolved over 300 generations using standard hyperparameters (crossover probability = 0.9, mutation probability = 0.25, and tournament size = 2). This process yielded a diverse ensemble of Pareto-efficient solutions balancing trade-offs between objectives. From the resulting Pareto front (@fig-pareto), we selected the top 30 parameter sets with the lowest combined error to use in the subsequent simulations assessing intervention effects. -->

To explore optimal parameter configurations, we used the NSGA-II algorithm as implemented in the `nsga2R` package [@nsga2R]. This algorithm is inspired by natural selection and evolves a population of candidate solutions over multiple generations to balance competing objectives. We began with a population of 500 randomly generated parameter sets and allowed them to evolve over 300 generations. During this process, solutions were selected, combined, and mutated using standard settings: a crossover probability of 0.9 (which controls how often solutions are combined), a mutation probability of 0.25 (introducing random variation), and a tournament size of 2 (used to pick the best candidates for reproduction).
The result is a Pareto front—a set of solutions that are considered optimal in the sense that no single objective can be improved without worsening another. From this diverse set of trade-off solutions (see @fig-pareto), we selected the 30 parameter sets with the lowest overall error across objectives. These top-performing configurations were then used in the simulation phase to evaluate the potential effects of different interventions.

```{r}
#| label: fig-pareto
#| fig-cap: >
#|   Pareto front of simulated parameter sets.
#|   *Each point represents a solution evaluated by its total variance error (x-axis) and total partial correlation error (y-axis), with color indicating total mean error.
#|   Red crosses highlight the top 30 solutions used for simulation.*
#| fig-align: center
#| out-width: 80%
knitr::include_graphics("img/paretoplot.png")
```

@@

To evaluate the effects of hypothetical support, we simulate scenarios in which each individual’s financial stress level ($S$) is proportionally reduced toward the minimum observed value. This models a population-wide intervention applied at varying strengths --- from no reduction in stress (0%) to complete alignment with the minimum observed stress level (100%).

For each of the top 30 parameter sets identified through calibration, we run simulations on 300 virtual individuals across this full range of intervention levels. In each simulation, we observe how depression ($D$) and precariousness ($P$) evolve over time, capturing their final values after the system stabilizes. This setup allows us to evaluate how different internal model dynamics (e.g., feedback strength) influence the system’s response to external change.

@@

To evaluate how internal system structure influences the effect of external support, we simulated outcomes under a range of intervention strengths across the top 30 parameter sets from the calibrated model ensemble. The top panel of @fig-intervention shows mean levels of depression ($D$) and precariousness ($P$) after simulation, with each line representing one parameter set. Lines are color-coded by combined feedback and input strength, calculated as the product $\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2$.

Parameter sets with stronger interaction dynamics exhibited more pronounced reductions in both depression and precariousness following intervention. At full intervention—modeled as shifting the external stressor ($S$) to its observed minimum—the reduction in depression ranged from [0.68*SD* – 0.97*SD*], and for social precariousness from [0.77*SD* – 0.94*SD*], corresponding to the lowest to highest values of parameter strength.[^3] These results suggest that systems with stronger internal dynamics are more responsive to external change, enabling targeted interventions to more effectively shift the system toward a lower-risk equilibrium.

[^3]: Standard deviations (*SD*) are based on the original scale of each variable: $\textit{SD}_D = 1.0$ for depression ($D$) and $\textit{SD}_P = 0.9$ for precariousness ($P$). Parameter strength is defined as the product $\alpha_1 \times \beta_1 \times \alpha_2 \times \beta_2$, reflecting the combined influence of internal feedback and external input sensitivity.

To better understand these dynamics, we focused on two illustrative parameter sets: one with strong feedback/input coupling and low noise (a *bistable* candidate), and one with weaker coupling and higher noise (a *unistable* candidate). In nonlinear systems, bistability refers to the existence of two distinct stable equilibria—meaning the system’s long-term outcome depends on its starting point or perturbations.

In our model, bistability arises from the interaction between strong internal coupling—particularly the product $\alpha_2 \times \beta_2$—and the system’s nonlinearity. The $\tanh(\cdot)$ activation function saturates at high input values, meaning sufficient gain is required to push the system into nonlinear regimes where multiple stable points emerge. If the feedback gain is too low, or the system is overwhelmed by noise, it behaves more linearly and settles into a single equilibrium—yielding a unistable configuration.[^1]

[^1]: This is similar to a well-known pattern in dynamical systems called a *pitchfork bifurcation*. In simple terms, when the system’s internal feedback becomes strong enough—for example, when $k$ in the equation $x_{t+1} = \tanh(k x_t)$ passes a certain threshold—the system changes from having one stable state to having two possible stable states (and one unstable one in between). In our model, this kind of shift happens when the feedback loop between depression and precariousness—captured by the product $\alpha_2 \times \beta_2$—becomes strong. It allows the system to settle into different long-term states, depending on initial conditions.

The bottom panel of @fig-intervention illustrates these contrasting behaviors. In the bistable model, a clear bifurcation emerges: small interventions have little effect, but once a threshold is crossed, the system abruptly transitions to a low-risk state. In the unistable model, the transition is gradual and continuous, with outcomes improving incrementally as stress is reduced.

These findings underscore that the impact of intervention depends not only on its intensity, but on the system's internal configuration. In bistable regimes, small or moderate interventions may be insufficient unless they push the system past a tipping point. Unistable systems, by contrast, respond more linearly, offering greater predictability and flexibility in designing incremental interventions.

This distinction is especially relevant for mental health interventions in populations exposed to persistent stressors like housing instability, financial strain, or job insecurity. In such contexts, external adversity may be reinforced through internal feedback loops between psychological distress and precarious conditions. Recognizing when a system is bistable may help identify when strong, early, or multifaceted interventions are needed to produce lasting change—while unistable dynamics may justify more gradual or targeted efforts.

Tailoring intervention strategies to the system’s underlying structure may enhance their effectiveness and sustainability, particularly in addressing the compounding effects of socioeconomic precariousness on mental health.

<!-- The final outcomes, specifically the mean levels of depression ($D$) and precariousness ($P$), are presented in the top panel of @fig-intervention. Each trajectory corresponds to a parameter set and is color-coded by its combined feedback and input strength, defined as the product $\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2$. -->

<!-- Systems with stronger parameter interactions exhibit more pronounced shifts in behavior, particularly under intensive interventions. At full intervention (i.e., maximum stress reduction), the mean depression level decreased from -0.113 to -0.963, and precariousness from -0.100 to -0.958. These changes represent reductions of approximately 90.8\% relative to their respective baseline magnitudes, indicating that effective intervention can push the system close to its lowest-risk equilibrium. -->

<!-- To explore these dynamics in more detail, we focused on two representative parameter sets: one with high feedback/input strength and low noise (a bistable candidate), and another with weak parameter interactions and high noise (a unistable candidate). In nonlinear dynamical systems, bistability refers to the existence of two distinct stable equilibria—meaning the long-term outcome depends on initial conditions or perturbations. -->

<!-- In our model, bistability emerges from an interplay between the strength of parameter-driven feedback/input loops and the magnitude of stochastic noise. The internal loop between $D$ and $P$ — governed by the product $\alpha_2 \times \beta_2$ — allows changes in one variable to reinforce the other. At the same time, the $\tanh(\cdot)$ nonlinearity in the system saturates for large inputs, flattening the system’s responsiveness. For multiple stable equilibria to emerge, the internal gain must be large enough to push the system into these nonlinear (saturated) regions. If this gain is too small, or if noise is too high, the system behaves more linearly and converges to a single stable point — a unistable regime.[^1]  -->

<!-- [^1]: This condition resembles the classical *pitchfork* bifurcation in discrete dynamical systems: when the gain $k$ in a scalar system such as $x_{t+1} = \tanh(k x_t)$ exceeds a critical threshold (e.g., $k>1$), the system admits three fixed points—two stable and one unstable—thus producing bistability. In our two-variable model, sufficiently high values of $\alpha_2 \times \beta_2$ similarly give rise to a self-reinforcing loop that sustains two attractor states. -->

<!-- This behavior is illustrated in the bottom panel of @fig-intervention. In the bistable model, we observe a bifurcation in final outcomes: as stress is reduced, the system undergoes a sharp transition from a high-$D$/$P$ regime to a low one—indicating a tipping point. In contrast, the unistable model shows a smooth, continuous response to increasing intervention, consistent with a single stable equilibrium. -->

<!-- These findings highlight how a system’s internal structure — including both nonlinear feedback gain and noise levels — fundamentally shapes its response to intervention. In bistable regimes, small or moderate interventions may have little effect unless they exceed a critical threshold sufficient to tip the system out of a high-risk state. In unistable systems, by contrast, outcomes improve more gradually and predictably with increasing support. -->

<!-- This distinction is particularly relevant when addressing persistent stressors associated with precariousness, such as unstable housing, employment insecurity, or chronic financial strain. In these contexts, the effects of external adversity may be reinforced internally through feedback loops between stress exposure and mental health. Such dynamics can stabilize harmful states and limit responsiveness to minor interventions—emphasizing the need for targeted, high-leverage support. -->

<!-- Recognizing whether a population’s mental health dynamics follow a bistable or unistable structure can guide policy design. For bistable systems, early or forceful interventions may be required to trigger lasting improvement. For unistable systems, more moderate or incremental measures may suffice. By tailoring strategies to the system’s underlying structure, interventions can be made more effective and context-sensitive—ultimately enhancing support for those facing precariousness-driven mental health risks. -->

```{r}
#| label: fig-intervention
#| fig-cap: "**Top**: Final Mean D and P Across Intervention Sizes. *Mean trajectories of depression ($D$) and precariousness ($P$) across varying intervention strengths, simulated over the top 30 parameter sets. Each line represents one parameter set and is colored by its combined feedback and input strength.* **Bottom**: Smoothed Density Heatmap of Final $D$ and $P$ for Two Parameter Sets. *One exhibits unistability (top two panels) and the other exhibits bistability (bottom two panels). Bistable dynamics reveal a sharp transition between distinct outcome regimes, while unistable dynamics show gradual, continuous change across the intervention gradient.*"
#| fig-align: center
#| out-width: 100%

knitr::include_graphics("img/intervention_plot2.pdf")
```

@@

To examine how such structures shape responsiveness to support, we developed a simplified two-variable nonlinear dynamical model capturing the feedback between depression and precariousness, modulated by financial stress. Calibrated against empirical summary statistics from the HELIUS dataset, the model allowed us to simulate how these dynamics evolve under varying levels of stress reduction.
The simulations revealed that intervention effectiveness depends not just on the magnitude of external stressors, but on the internal configuration of the system itself. In bistable regimes—characterized by strong feedback and low noise—the system could remain stuck in a high-risk state unless an intervention crosses a critical threshold, tipping it into a healthier equilibrium. By contrast, unistable systems—driven by weaker coupling or greater noise—responded more gradually and proportionally to increasing support.
These dynamics carry practical policy implications. In bistable systems, even moderate interventions may have little effect unless they are forceful enough to tip the system toward a lower-risk state. This underscores the importance of early or intensive support when targeting self-reinforcing cycles of depression and precariousness. Unistable systems, by contrast, are more responsive to incremental support, suggesting that sustained, moderate interventions may be sufficient in such contexts.

<!-- <!-- Interpretation of Findings --> 
<!-- Our findings highlight the significant role of recent financial stress (S.fin) as a potential causal factor for depression, consistently observed across both sum score and symptom-level analyses. This consistency underscores the profound impact of financial stressors on mental health. The symptom-level analysis further identified specific symptoms—particularly sleep disturbances (slp), depressed mood (dep), guilt (glt), and anhedonia (anh)—as being especially sensitive to external precarious conditions. -->

<!-- Among these symptoms, dep stands out as a distinct causal factor for precariousness in both aggregated and disaggregated analyses, with CCI even suggesting a possible cyclic relationship between them. This implies that depression and precariousness may reinforce one another, potentially leading to a worsening cycle over time. Within the symptom network, dep also exhibits a high degree of connectivity, but with mostly incoming arrows, suggesting that as symptom dynamics intensify, dep is quickly impacted by multiple factors—particularly P.soc (social precariousness), further fueling the depression-precariousness cycle. -->

<!-- Beyond dep, slp emerges as a critical symptom in this dynamic. It connects with both social and employment precariousness, while also playing a central role in symptom interactions, affecting other symptoms more than most others. This suggests that slp may act as an initiator within the depressive symptom system, potentially triggering or amplifying symptom cascades. -->

<!-- Among the precariousness factors, P.soc appears to be one of the most pivotal—not as a driver of symptoms, but rather as a consequence of depression. It maintains the strongest connectivity with symptoms, yet its directionality indicates that it is largely shaped by depressive symptoms rather than acting as a primary cause. Together, these patterns suggest a broader directional dynamic in which financial stress (S.fin) exacerbates depressive symptoms, which in turn feed into social precariousness, perpetuating the cycle. -->

<!-- This directional insight suggests that interventions targeting specific depressive symptoms, particularly dep, may help prevent downstream effects on worsening precarious conditions. Alternatively, early intervention focusing on slp may be a promising strategy, as slp appears to act as a potential initiator or activator within the depressive symptom network. This positioning suggests that slp could serve as an early warning sign, offering a valuable target for preventive interventions aimed at breaking the depression-precariousness cycle before it intensifies. -->

<!-- Building on these insights, we developed a simplified computational model capturing the core mechanisms identified in the data: bidirectional feedback between depression and precariousness, driven by external stress. Calibrated against empirical summary statistics, the model allowed us to explore how different parameter configurations—particularly feedback strength and stochastic noise—modulate the system’s response to simulated interventions. Some systems exhibited bistability, resisting change until pushed past a critical threshold, while others responded more gradually and predictably. These patterns have direct policy implications: in bistable regimes, intensive or early interventions may be required to shift individuals out of high-risk states, whereas in unistable systems, more moderate, sustained support may suffice. -->

<!-- A key contribution of this work is to show that the structure of the depression–precariousness system matters: not all forms of adversity respond equally to intervention, and this heterogeneity may stem from internal feedback dynamics as much as from external conditions. Recognizing whether a given population operates within a bistable or unistable regime could inform the design and timing of mental health interventions—particularly in contexts of chronic socioeconomic stress. -->

<!-- Building on these findings, we developed a simplified two-variable nonlinear dynamical system to simulate the co-evolution of depression and social precariousness under varying levels of financial stress. This model served as a conceptual abstraction of the inferred causal structure, enabling us to assess how the system’s internal configuration—particularly feedback strength, input coupling, and stochastic noise—modulates its behavior in response to intervention. -->

>>>>

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    template-partials:
      - title.tex
    include-in-header: header.tex
    crossref:
      appendix-title: "App."
      appendix-delim: ":"
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
nocite: |
  @*
csl: apa.csl
abstract: |
  \noindent This study investigates how the internal structure of depression–precariousness dynamics shapes responsiveness to external support. Using cross-sectional data from the HELIUS study, we first apply cycle-capable causal discovery algorithms to estimate directional dependencies between financial stress, domain-specific precariousness, and depressive symptoms. Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—highlights financial stress as a plausible upstream driver of depression, with symptoms such as sleep disturbance and depressed mood emerging as initiators of broader social vulnerability.
  To translate these structural insights into dynamic predictions, we construct a family of analytically calibrated linear stochastic models. Simulations reveal that variations in internal structure can substantially alter system responses to financial stress reduction. Some models show rapid but transient improvements in depression, while others adjust more slowly yet sustain gains over time. Precariousness tends to respond more immediately across all configurations, though its trajectory also shifts subtly with structural variation.
  Together, these findings demonstrate that intervention outcomes depend not only on the magnitude of external input but also on latent properties of the system. More broadly, this study illustrates how combining causal discovery with dynamical modeling supports principled, simulation-based exploration of intervention scenarios—grounded in data, yet able to probe counterfactual structure. This approach offers a scalable framework for diagnosing hidden system constraints and informing targeted public health strategies in complex psychosocial domains.

@@

As illustrated in @fig-concept, the model includes three components: depression ($D$), represented by the PHQ-9 sum score, social precariousness ($P$), and external financial stress ($S$). Depression and precariousness are modeled as coupled state variables that influence each other over time, with $S$ acting as a shared input. These dynamics are formalized as a pair of linear stochastic differential equations (SDEs):

$$
\begin{aligned}
dP &= \lambda_P \left(\alpha_{PS} S + \alpha_{PD} D - P \right) dt + \sigma_{P} dW_{P} \\
dD &= \lambda_D \left(\alpha_{DS} S + \alpha_{DP} P - D \right) dt + \sigma_{D} dW_{D} 
\end{aligned}
$$

Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress. The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which each variable adjusts toward its input-driven target state. The coefficients $\alpha_{PS}$ and $\alpha_{DS}$ quantify the direct effects of the stressor on $P$ and $D$, while $\alpha_{DP}$ and $\alpha_{PD}$ represent the cross-influences between depression and precariousness. Stochastic fluctuations are introduced through the noise terms $\sigma_P, dW_P$ and $\sigma_D, dW_D$, with $dW_P$ and $dW_D$ representing independent Wiener processes.
This linear structure offers a transparent and analytically tractable framework for examining how variations in external financial stress influence the population-level dynamics of depression and social precariousness. Crucially, the model's simplicity enables direct calibration from empirical data: all parameters can be derived analytically from observed variances and covariances in the HELIUS cohort. The following  [@sec-calibration] details the procedure used to perform this calibration.

@@

### Model Calibration {#sec-calibration}

In our linear model, all parameters are derived analytically to reproduce the empirical second-order moments observed in the HELIUS dataset. The system treats depression ($D$) and social precariousness ($P$) as jointly evolving under the influence of a shared external stressor ($S$), forming a coupled system of Ornstein–Uhlenbeck processes. Under the assumption of stationarity, the steady-state covariances between these variables admit closed-form solutions for the system’s structural parameters.

We fix the adjustment rates $\lambda_D = \lambda_P = 1$, which normalizes the time scale given the cross-sectional nature of the HELIUS data.
To resolve the remaining degrees of freedom, we treat the feedback from precariousness to depression ($\alpha_{DP}$) as a free parameter, and solve algebraically for all other coefficients — including the direct effects of stress ($\alpha_{DS}, \alpha_{PS}$), the reverse feedback from depression to precariousness ($\alpha_{PD}$), and the noise terms ($\sigma_D, \sigma_P$). These solutions are expressed as closed-form functions of the empirical covariances: $\mathrm{Var}(D)$, $\mathrm{Var}(P)$, $\mathrm{Var}(S)$, $\mathrm{Cov}(D, P)$, $\mathrm{Cov}(D, S)$, and $\mathrm{Cov}(P, S)$.

This structure creates a family of analytically calibrated models, each consistent with the same observed covariances but differing in their internal configuration. However, we constrain the space of valid solutions by imposing two empirical and theoretical conditions: (1) all structural parameters must be positive, reflecting the assumption that stress and precariousness contribute adversely to depression; and (2) the system must be dynamically stable, ensuring convergence to a stationary distribution. These constraints restrict the allowable range of $\alpha_{DP}$ to approximately $[0, 0.698]$ (see Appendix [@sec-analcalibration]).

@fig-tradeoff illustrates how the remaining parameters adjust as $\alpha_{DP}$ increases to maintain consistency with the empirical covariances. As $\alpha_{DP}$ increases, $\alpha_{PD}$ must decline to preserve stability, since the product $\alpha_{DP} \cdot \alpha_{PD}$ determines the strength of the feedback loop and must remain below 1. This creates a compensatory trade-off: as influence from $P$ to $D$ strengthens, the influence from $D$ to $P$ weakens. Meanwhile, the external coupling terms — $\alpha_{DS}$ and $\alpha_{PS}$ — also shift to absorb variance in a way that maintains alignment with the empirical data. Notably, the noise amplitudes $\sigma_D$ and $\sigma_P$ decline slightly as $\alpha_{DP}$ increases. This indicates that more of the observed variability is captured by deterministic interactions, with less left to be explained by stochastic fluctuations.

Rather than being a limitation, this flexibility enables exploration of a spectrum of equally data-consistent feedback regimes. Each represents a different plausible causal architecture, offering insight into how internal dynamics modulate the system's responsiveness to external stress — even when the observed covariances remain unchanged.

```{r}
#| label: fig-tradeoff
#| fig-cap: >
#|   Trade-off curves showing how calibrated parameters vary with $\alpha_{DP}$.
#|   *Each line shows how a different parameter adjusts in response to changes in $\alpha_{DP}$ to maintain consistency with empirical covariances from the HELIUS data. 
#|   All configurations shown are equally consistent with the data, illustrating how internal dynamics adjust as feedback strength changes.*
#| fig-align: center
#| out-width: 100%
knitr::include_graphics("img/param_tradeoff.pdf")
```

@@

To evaluate how internal structure affects system responsiveness, we simulate a family of linear models subjected to external intervention. Each model corresponds to a different value of $\alpha_{DP}$, representing the strength of influence from social precariousness to depression. The range spans from 0 (no coupling) to 0.65 (maximal admissible value), with all other parameters derived analytically to preserve the empirical covariances observed in the HELIUS data (see @fig-tradeoff).

In each simulation, financial stress ($S$) is proportionally reduced toward the lowest observed value, modeling a population-wide intervention applied at varying intensities (0–100%). For each model, we simulate 1000 virtual individuals whose baseline stress levels are sampled from the empirical distribution and scaled according to the intervention level.

We assess the system’s behavior using two complementary approaches:

- Snapshot analysis captures the average levels of depression ($D$) and precariousness ($P$) at selected time points ($t = 100$, $300$, $500$) following intervention onset. Since all linear models converge to the same equilibrium, this highlights how internal feedback affects the speed and pathway of adjustment.

- Full trajectory simulation tracks system dynamics over time for two representative models: one with minimal ($\alpha_{DP} = 0$) and one with maximal ($\alpha_{DP} = 0.65$) feedback. Each simulation includes an intervention period followed by its withdrawal, revealing differences in reactivity, delay, and recovery stability.

Together, these simulations demonstrate how internal structure governs not just whether change occurs, but how it unfolds — shaping the pace, persistence, and variability of response to external support.

@@

To assess how internal structure influences the system’s responsiveness to external support, we simulate a set of linear models that differ in the assumed strength of influence from precariousness to depression, $\alpha_{DP}$. For each model, we generate a virtual population of 1,000 individuals and examine system behavior in two complementary ways: by analyzing average outcomes at selected time points following intervention onset, and by simulating full temporal trajectories for two representative cases at the extremes of the $\alpha_{DP}$ range.

@fig-intervention summarizes how the internal structure of the model—specifically, variation in $\alpha_{DP}$—shapes the system’s short-term and long-term responses to external support. Panel A shows the mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500 after intervention onset, across a range of $\alpha_{DP}$ values. Each line represents a different model, color-coded from low (purple) to high (yellow) values of $\alpha_{DP}$.

In the early phase of intervention ($t = 100$), depression and precariousness exhibit contrasting patterns. Depression declines more rapidly in models with lower $\alpha_{DP}$, where its dynamics are driven more directly by the external stressor $S$. As $\alpha_{DP}$ increases, $D$ becomes more dependent on precariousness ($P$), which itself adjusts more slowly—delaying the depressive response.
Precariousness, by contrast, shows slightly faster early reductions in models with higher $\alpha_{DP}$. This reflects a shift in influence: as $\alpha_{DP}$ increases, $P$ becomes more directly responsive to $S$ (via increasing $\alpha_{PS}$) and less influenced by $D$, allowing it to adjust more promptly to changes in the stressor.

Despite these short-term differences, all models ultimately converge to similar steady-state levels under sustained intervention. By time step 300, convergence is already substantial—though some variation remains, particularly in $P$, where gradients between models are still visible. By step 500, trajectories have nearly fully aligned across all configurations, consistent with the shared equilibrium structure of the linear system. This convergence highlights that the key distinctions lie not in whether intervention is effective, but in how quickly and through which pathways change unfolds.

Panel B shows full time trajectories for two representative models: one with no coupling ($\alpha_{DP} = 0$) and one at the upper admissible limit ($\alpha_{DP} = 0.65$). In both, intervention begins at step 300 and ends at step 1500. In the uncoupled model, depression responds rapidly and rebounds shortly after support is removed. In the high-$\alpha_{DP}$ model, $D$ responds more gradually and recovers more slowly—reflecting its greater dependence on the slower-changing $P$.
Precariousness declines in both models during intervention and returns to baseline afterward, but with a slightly sharper drop in the higher $\alpha_{DP}$ condition. This is because $P$ becomes more directly sensitive to $S$ as $\alpha_{PS}$ increases with $\alpha_{DP}$.

These asymmetries further highlight how internal structure—not just external inputs—shapes the system’s trajectory. Specifically, changes in coupling parameters reconfigure how stress reduction propagates through the system, altering the speed and pathway of response even when long-run outcomes remain unchanged.

```{r}
#| label: fig-intervention
#| fig-cap: >
#|   System response to external intervention across variations in internal structure.
#|   ***Panel A**: Mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, across different values of $\alpha_{DP}$ and intervention strengths.
#|   Models with lower $\alpha_{DP}$ show faster early reductions in $D$, while higher $\alpha_{DP}$ leads to more immediate declines in $P$.
#|   **Panel B**: Full time trajectories for two models—one with no coupling ($\alpha_{DP} = 0$) and one with strong coupling ($\alpha_{DP} = 0.65$)—illustrating system response and recovery following intervention onset (step 300) and withdrawal (step 1500).
#|   Depression responds more gradually and persistently in the high $\alpha_{DP}$ model, whereas precariousness adjusts more quickly.*

#| fig-align: center
#| out-width: 100%

knitr::include_graphics("img/intervention_linear.pdf")
```

@@

To explore how internal structure shapes response to external support, we developed a linear stochastic model of depression and precariousness, influenced by financial stress. Crucially, we analytically calibrated this model to match empirical covariances from the HELIUS data, ensuring that all parameter sets were observationally equivalent but differed in their internal configuration. We treated the parameter $\alpha_{DP}$—the effect of precariousness on depression—as a free variable, and derived all other parameters accordingly, yielding a family of valid models consistent with the data.
Simulation results showed that even with identical steady-state covariances, differences in internal structure produced systematic variation in the timing and profile of system responses to intervention. In models with lower $\alpha_{DP}$, depression responded more quickly to reduced stress because its dynamics were more directly coupled to the external input. In contrast, as $\alpha_{DP}$ increased and depression became more indirectly influenced—mediated by the slower-changing precariousness variable—its response was more gradual and persistent. Precariousness, while generally more directly responsive across all models, also showed subtle differences, with slightly faster early declines in configurations where it was more sensitive to external stress.
These results demonstrate that intervention impact depends not only on the external conditions (e.g., stress reduction), but also on the internal structure of the system—how variables are coupled and through which pathways change flows. Tailoring interventions to system configuration—using more intensive or earlier support in systems with indirect or delayed pathways, and sustained support in systems with rapid but unstable responses—may help optimize mental health outcomes under conditions of social and economic adversity.

<!-- To examine how such structures shape responsiveness to support, we developed a simplified two-variable nonlinear dynamical model capturing the feedback between depression and precariousness, modulated by financial stress. Calibrated against empirical summary statistics from the HELIUS dataset, the model allowed us to simulate how these dynamics evolve under varying levels of stress reduction. -->
<!-- The simulations revealed that intervention effectiveness depends not just on the magnitude of external stressors, but on the internal configuration of the system itself. In bistable regimes—characterized by strong feedback and low noise—the system could remain stuck in a high-risk state unless an intervention crosses a critical threshold, tipping it into a healthier equilibrium. By contrast, unistable systems—driven by weaker coupling or greater noise—responded more gradually and proportionally to increasing support. -->
<!-- These dynamics carry practical policy implications. In bistable systems, even moderate interventions may have little effect unless they are forceful enough to tip the system toward a lower-risk state. This underscores the importance of early or intensive support when targeting self-reinforcing cycles of depression and precariousness. Unistable systems, by contrast, are more responsive to incremental support, suggesting that sustained, moderate interventions may be sufficient in such contexts. -->
